\section{Related Work}\label{sec:related}

The development of quantum programming languages and frameworks has evolved significantly over the past two decades, with researchers proposing various approaches to address the challenges of quantum-classical integration, abstraction levels, and error handling. This section systematically reviews significant contributions to quantum programming models, focusing particularly on their approaches to managing the hybrid nature of practical quantum computation, which is central to our work.

\subsection{Quantum Programming Languages}

The evolution of quantum programming languages reflects the ongoing effort to balance theoretical rigor with practical implementation considerations. We categorize these languages based on their design philosophies and implementation strategies.

\subsubsection{Foundational Languages}

Several languages have established the theoretical foundations for quantum programming, focusing on formal semantics and mathematical rigor.

\textbf{Quantum Programming Language (QPL)}~\cite{selinger2004towards}, introduced by Peter Selinger in 2004, was one of the first formal languages specifically designed for quantum computing. QPL features a rigorous semantics based on quantum operations and completely positive maps, enabling formal verification of quantum programs. It introduced a flow chart syntax for representing quantum programs, with explicit typing for both classical and quantum data. QPL's most significant contribution was its denotational semantics using superoperators on density matrices, which became a foundation for subsequent formal approaches to quantum programming languages. While QPL established important theoretical principles for quantum language design, it was primarily a formal system without practical implementations targeting physical quantum hardware platforms.

\textbf{Quipper}~\cite{green2013quipper}, developed by Green et al., is a scalable functional programming language for quantum computing embedded in Haskell. Quipper leverages Haskell's advanced type system and higher-order functions to enable powerful circuit generation and manipulation capabilities. It introduces quantum data types, gate operations, and circuit transformations as first-class objects, allowing for sophisticated metaprogramming techniques like circuit specialization and oracle synthesis. Quipper is designed for a model of computation where a classical computer controls a quantum device, and has demonstrated the ability to generate quantum circuits with trillions of gates. It pioneered the concept of circuit families - parameterized descriptions that generate specific quantum circuits based on input parameters - making it particularly valuable for expressing algorithms with varying problem sizes. While powerful for algorithm expression, Quipper's integration with classical control relies primarily on Haskell's host language mechanisms rather than intrinsic quantum language features, creating a separation between quantum and classical components that differs from our integrated approach.

\textbf{Twist}~\cite{yuan2022twist}, introduced by Yuan et al. at MIT, formalizes quantum state purity as a central tool for automating reasoning about entanglement in quantum programs. Twist's key innovation is its type system for sound reasoning about purity in quantum programs, where a pure expression is one whose evaluation is unaffected by measurement outcomes of qubits it does not own, implying freedom from entanglement with other expressions in the computation. The language enables developers to identify pure expressions using type annotations and features purity assertion operators that state the absence of entanglement in the output of quantum gates. To soundly check these assertions, Twist uses a combination of static analysis and runtime verification. Evaluation of Twist's approach on quantum algorithm benchmarks demonstrated that it can express quantum algorithms, catch programming errors, support programs that existing languages disallow, while incurring reasonable runtime verification overhead. While primarily focused on formal verification rather than general-purpose programming, Twist's rigorous approach to reasoning about entanglement provides important theoretical foundations for safe quantum programming language design.

\subsubsection{Quantum Assembly Languages}

Quantum assembly languages serve as important intermediaries between high-level quantum programming and hardware execution, providing a balance between hardware-agnostic programming and efficient execution. These languages focus on defining universal instruction sets that can serve as common compilation targets for higher-level languages.

\textbf{eQASM}~\cite{fu2019eqasm} (executable QASM), designed specifically for superconducting quantum processors, introduces a two-level instruction decoding scheme that separates quantum operations from their timing specifications. It includes microcode support and a comprehensive quantum instruction set architecture (QuISA) that bridges the gap between high-level algorithms and physical control pulses. eQASM provides explicit support for fast feedback through its Active Reset and Fast Conditional Operations, addressing the challenge of real-time classical-quantum interaction, though its design is closely tied to superconducting qubit architectures.

\textbf{cQASM}~\cite{khammassi2018cqasm} (common Quantum Assembly Language) was developed at QuTech to serve as a standardized language for representing quantum circuits across different platforms and simulators. It aims to abstract away qubit technology details and guarantee interoperability between quantum compilation and simulation tools. cQASM 1.0 focused on establishing a common syntax for quantum circuit description to overcome the fragmentation caused by multiple QASM dialects, which had created compatibility issues between different quantum computing tools. The language was designed to facilitate sharing of quantum programs across the research community without requiring translation between incompatible syntaxes. Like many quantum assembly languages, cQASM follows the established convention where classical instructions control quantum operations.

\textbf{QUASAR}~\cite{butko2020understanding} is a scalar quantum instruction set architecture proposed as an extension to RISC-V for quantum control processors. Designed specifically to address the challenges of intermediate-scale quantum devices, QUASAR provides functional abstractions that map high-level semantics of quantum programming languages to low-level pulse generation by hardware. It focuses on circuit encoding efficiency and meeting the real-time gate cycle requirements of quantum chips. The architecture was evaluated against metrics such as performance, timing constraint implications, and scalability for future quantum processors, providing important design considerations for quantum control systems that must balance computational capabilities with the stringent timing requirements of quantum operations.

These quantum assembly languages generally aim to provide hardware-independent abstractions while remaining close enough to physical operations to enable efficient execution. However, they often struggle with the challenge of seamlessly integrating classical and quantum computation, particularly for dynamic quantum algorithms requiring adaptive measurements and feedback. Our approach differs by addressing the classical-quantum boundary at a more fundamental level through our stack-based execution model.

\subsubsection{Hardware-Oriented Languages}

Hardware-oriented quantum programming languages focus on providing abstractions that closely match the capabilities of existing or near-term quantum computing hardware, often prioritizing efficiency and direct mapping to physical operations.

\textbf{OpenQASM}~\cite{cross2017open,cross2022openqasm} has emerged as one of the most widely adopted quantum assembly languages, particularly within the IBM Quantum ecosystem. Starting with OpenQASM 2.0's simple gate-based description language, it has evolved significantly with OpenQASM 3.0 introducing a comprehensive programming model that includes classical control flow, user-defined gates, timing constraints, and multi-level programming abstractions. OpenQASM 3.0 supports real-time classical computation through its "let" statements and feed-forward operations with "if" statements, allowing runtime decisions based on measurement results. Despite these improvements, OpenQASM follows the QRAM paradigm where classical control drives quantum operations, which can create challenges when implementing error correction protocols that require intricate feedback between classical and quantum components.

\textbf{Quil}~\cite{smith2016practical}, developed by Rigetti Computing, is a quantum instruction set architecture designed to support hybrid quantum-classical computation. The original Quil specification provided a gate-level representation for quantum circuits with basic classical control capabilities. Quil-T, its temporal extension, introduced significant enhancements for hybrid quantum-classical computing, including more sophisticated classical control flow, timing specifications, and multi-qubit gate decomposition rules. A distinguishing feature of Quil is its explicit ``classical memory regions'' model, which defines typed memory spaces that can be read from and written to by both quantum measurements and classical instructions. This approach facilitates direct communication between quantum and classical processors but differs from our stack-based approach, as it relies on explicitly named registers that must be statically allocated rather than dynamically managed through our nested kernel structure.

\subsubsection{High-Level Programming Languages}

High-level quantum programming languages aim to provide abstractions that make quantum algorithm development more accessible to programmers without expertise in quantum physics, while maintaining the ability to express complex quantum operations efficiently.

\textbf{Q\#}~\cite{qsharp}, developed by Microsoft's Quantum group, represents one of the most comprehensive attempts to create a high-level quantum programming language integrated with classical control. Q\# implements a hybrid quantum-classical programming model with strong type safety, including specialized quantum data types (qubits, operations, and results) and operations as first-class program elements. It features a unique capability-based qubit management system that enforces safe qubit usage patterns. The language supports complex quantum-classical interactions through its native control flow constructs (conditionals, loops) and measurement feedback capabilities. Q\# programs operate under a QRAM execution model where host-driven classical code controls quantum operations, contrasting with our approach where quantum execution is primary and classical operations are integrated as callbacks, reflecting different design priorities in managing quantum resources and control flow.

\textbf{Silq}~\cite{silq}, developed at ETH Zurich, represents a significant advancement in high-level quantum programming languages. Its key innovation is automatic uncomputation of temporary quantum values, relieving programmers of the complex task of manually ensuring quantum garbage collection through reverse operations. Silq introduces a novel type system that distinguishes between classical and quantum data, and tracks whether quantum data is consumed during operations. It also features a more intuitive programming model with natural quantum control flow and safer semantics for quantum measurements. Unlike our approach, Silq handles the quantum-classical boundary primarily through a compile-time transformation process rather than a runtime integration model, reflecting a different philosophy in managing the complexity of quantum resource manipulation.

\subsection{Quantum Simulation and Error Correction Frameworks}

The development of specialized frameworks for quantum simulation and error correction has been critical in bridging the gap between theoretical quantum algorithms and their practical implementation on error-prone quantum hardware.

\textbf{LIQUi}$|\rangle$~\cite{wecker2014liqui}, developed by Microsoft's Quantum Architectures and Computation Group (QuArC), is a comprehensive software architecture and toolsuite for quantum computing. The name stands for Language-Integrated Quantum Operations, and it provides a full-stack environment for quantum algorithm design, simulation, and execution. LIQUi$|\rangle$ includes a domain-specific language embedded in F\# that allows the high-level expression of quantum algorithms, along with compilers, optimizers, and simulators capable of handling up to 30 qubits on standard hardware. A distinguishing feature is its modularity, which permits the representation of quantum circuits as data structures that can be manipulated, optimized, and transformed across different abstraction levels. LIQUi$|\rangle$ supports a variety of quantum algorithms including quantum teleportation, Shor's factoring algorithm, quantum chemistry simulations, and quantum error correction. Its approach to managing the quantum-classical interface through a high-level functional programming model shares conceptual similarities with our stack-based approach, though they differ in implementation details.


\textbf{Stim}~\cite{gidney2021stim} is a highly specialized framework designed specifically for simulating stabilizer circuits and quantum error correction protocols with exceptional efficiency. Developed by Craig Gidney at Google Quantum AI, it can simulate Clifford+measurement circuits with tens of thousands of qubits through its optimized tableau representation and noise modeling capabilities. Stim includes explicit support for classical feedback channels and detector-based error analysis, making it particularly valuable for analyzing and validating error correction codes. Its detection event sampling and analysis tools enable precise characterization of error correction performance under realistic noise conditions. While Stim's focus on stabilizer circuits and error correction aligns conceptually with aspects of our work, it implements this through specialized simulation rather than a general-purpose programming model.


\subsection{Classical-Quantum Integration Models}

The integration of classical and quantum computation represents one of the most significant challenges in quantum programming language design. Current approaches to addressing this challenge generally fall into three distinct categories, each with its own advantages and limitations:

\begin{enumerate}
\item \textbf{Host language embedding}: Systems like Qiskit~\cite{Qiskit}, Cirq~\cite{cirq}, and PyQuil~\cite{pyquil} embed quantum operations within classical host languages (typically Python), with classical control handled by the host language's execution model. While this approach offers flexibility and familiarity to classical programmers, it often obscures the runtime requirements of quantum-classical interactions and can make it difficult to reason about the behavior of complex quantum programs, particularly those involving error correction or adaptive measurement.

\item \textbf{Extended circuit models}: Languages like OpenQASM 3.0 and Quil extend circuit descriptions with classical control and feedback mechanisms to support hybrid computation. While this approach maintains the circuit-based model familiar to quantum algorithm designers, it often provides limited expressiveness compared to general-purpose programming languages when implementing complex classical processing of measurement results. This limitation becomes particularly apparent when implementing sophisticated error correction protocols that require non-trivial classical decoding.

\item \textbf{Specialized hybrid languages}: Systems like Q\# provide integrated languages with both classical and quantum constructs, unifying the programming model and providing strong static guarantees. However, these languages still lack enough expressiveness to implemented complex classical decoding like Minimum Weight Perfect Matching \cite{higgott2022pymatching} or Belief Propagation\cite{yao2024belief}.
\end{enumerate}

Our stack-based approach introduces a fundamentally different model by inverting the traditional relationship between classical and quantum components. Rather than viewing quantum operations as subroutines called from classical code, qstack treats the quantum program as primary, with classical operations integrated as callbacks when measurement outcomes need processing. This inversion of control significantly simplifies the expression of quantum algorithms that require intricate patterns of measurement and classical feedback, such as error correction protocols and adaptive quantum algorithms. By making the quantum-classical boundary explicit through our kernel-based model, we also provide clearer reasoning about resource usage and program behavior.