\section{Quantum Instruction Generation}\label{sec:synthesis}

Generating quantum instructions from a fixed instruction set that compose to perform a computation is a core pattern available in most quantum programming languages. This translation between different levels of abstraction and different quantum instruction sets is fundamental to bridging the gap between algorithm design and hardware execution. This functionality exists in many frameworks like Qiskit, Cirq, and Q\#, and it is also crucial in qstack, where we have a consistent programming model across quantum-instruction abstraction levels.

The previous examples used qstack's toy instruction set with intuitive names like \texttt{mix}, \texttt{entangle}, and \texttt{skew} to make quantum operations more accessible. However, quantum programs are typically expressed in terms of a \textit{canonical} instruction set consisting of commonly used quantum gates with standardized names. These gates, also called quantum instructions, are unitary operations that transform quantum states while preserving quantum properties like superposition and entanglement. The canonical set usually includes Clifford operations --- the Pauli gates ($X$, $Y$, $Z$), Hadamard ($H$), phase ($S$), and controlled-X ($CX$)—along with arbitrary single- or two-qubit rotations. This set is universal, meaning it can approximate any quantum operation to arbitrary precision when gates are combined appropriately.

In practice, quantum hardware does not necessarily implement this canonical set natively. Instead, each quantum processor provides its own minimal universal instruction set based on its physical capabilities and constraints. Since all universal sets can express the same operations, they are equivalent up to a change of basis and can be transformed into one another through compilation, a process that replaces high-level gates with equivalent sequences of hardware-native gates.

As an example, take Quantinuum's Trapped-Ion quantum processor\cite{moses2023race}, which exposes as its universal instruction set two one-qubit operations $U_1$ and $RZ$, and the Mølmer-Sørenson-Gate ($RZZ$) as its entanglement operation. With these instructions, the Hadamard operation can be synthesized as:

$$H =  RZ(\pi) \circ U_1(\pi/2, -\pi/2) $$

and the two qubit Controlled-X as:

$$CX = I \otimes RZ(-\pi/2) \circ I \otimes U_1(\pi/2, \pi) \circ RZ(-\pi/2) \otimes I \circ RZZ(\pi/2) \circ I \otimes U_1(-\pi/2, \pi/2)$$

For these instructions, the complete compilation chain shows how the same Bell state program can be expressed at three different levels. Starting from qstack's toy instruction set, compiling to the canonical instruction set, and then compiling to Quantinuum's hardware-specific instruction set:

\begin{minted}[fontsize=\footnotesize]{ruby}
# Toy instruction set (intuitive)
allocate q1 q2:
  mix q1
  entangle q1 q2
measure

# Canonical instruction set (standard)
allocate q1 q2:
  h q1
  cx q1 q2
measure

# Quantinuum instruction set (hardware-specific)
allocate q1 q2:
  u_1(pi/2, -pi/2) q1
  rz(pi) q1
  
  u_1(-pi/2, pi/2) q2
  rzz(pi/2) q1 q2
  rz(-pi/2) q1
  u_1(pi/2, pi) q2
  rz(-pi/2) q2
measure
\end{minted}

The main point is that translating among instruction sets does not require leaving the qstack programming model. The compiled program remains a valid qstack program that can be further compiled, analyzed, or executed using the same tools and semantics. This composition enables modular compilation pipelines where different compilers can be chained together, each targeting specific aspects like instruction set translation, optimization, or error correction.

An important aspect of quantum circuit synthesis is the handling of non-Clifford operations, such as arbitrary rotations and T-gates, which are essential for universal quantum computation but typically more resource-intensive to implement in fault-tolerant settings. Modern synthesis techniques, like Solovay-Kitaev decomposition or more advanced methods using phase polynomials and trigonometric approximations~\cite{dawson2005solovay,kliuchnikov2013asymptotically}, approximate these operations to arbitrary precision using sequences of simpler gates. In qstack, these synthesis techniques can be integrated as specialized compiler passes that not only translate non-Clifford operations into hardware-native instruction sequences but also seamlessly interact with the callback-based feedback mechanisms demonstrated in our T-gate example. This synergy is particularly valuable when implementing fault-tolerant non-Clifford operations, where the magic state injection pattern (with its measurement and conditional correction logic) can be automatically inserted by a compiler pass while preserving the algorithm's high-level structure.
